A Plasma Torus Around a Young Low-Mass Star

### 2025-05-28 paper/ms.tex
update

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.99\textwidth]{f3.pdf}
  \vspace{-0.3cm}
  \captionsetup{labelformat=moviefmt,labelsep=colon}
	\caption{\textbf{Emission from circumstellar plasma orbiting
TIC\,141146667.}
  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{{\bf
  available here}},
  shows the spectral evolution over five hours.
  {\bf a,} Average TESS light curve from 5 February 2024 to 26
  February 2024 folded on the \periodhr\ hour period.  Black
  points are phase-averaged; gray are the raw data.
  {\bf b,} Keck/HIRES H$\alpha$ spectra acquired on 17 February 2024.
  The continuum is set to unity, and the darkest color is at twice the
  continuum to accentuate emission outside the line core ($|v/v_{\rm
  eq}|>1$, for the equatorial velocity $v_{\rm eq}$=130\,\kms).  While
  emission in the line core comes from the star's chromosphere, the
  sinusoidal emission features are most readily described by a warped
  plasma torus.
  {\bf c,} Individual epochs of Panel b, visible in the online movie.
  The dotted line shows the time-averaged spectrum, $f_{\langle t
  \rangle}$.
  {\bf d,} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf e,} As in Panel b, after subtracting the time-averaged
  spectrum.  The line core shows H$\alpha$ excesses and decrements
  advancing from the blue to red wings.
  The asymmetric color stretch
  is set to mirror the dynamic range of the data.
  {\bf f,} Individual epochs of Panel e, visible in the online
  movie.}
  \label{fig:spec}
\end{figure*}

@@

The H$\alpha$ line core\added{, visible in Panels e and f of
Figure~\ref{fig:spec}, }is more complex.  At $|\Delta v / v_{\rm
eq}|<1$, most observed H$\alpha$ photons come from the star's
chromosphere; \added{variability in the stellar emission line can be
caused by circumstellar material, bright regions, or dark regions
crossing the star's surface}.  Figure~\ref{fig:spec}e suggests that
\added{all three effects occur in TIC\,141146667}.  For instance, from
$\phi$=0-0.3, double-peaked emission is visible both on and off-limb;
this feature is circumstellar in origin.  However, the emission
feature that crosses the star from $\phi$=0.4-0.9 is brighter than the
circumstellar clumps, and its slow speed instead suggests it is
associated with bright region on the star's surface.  Similarly, from
$\phi$=0.6-1.15 a 20\% deep absorption feature slowly crosses the
H$\alpha$ line profile.  This feature could be either a
chromospherically dark region (e.g.~a spot group), or an azimuthally
extended component of the circumstellar material.  The origin of the
other bright and dark streaks passing across the line core \added{in
Figure~\ref{fig:spec}e }are similarly ambiguous.

@@

Spectra of magnetically-active, rapidly rotating stars with a wide
range of masses have been previously observed to exhibit sinusoidal
time-varying Balmer emission
\citep{Donati2000,Townsend2005b,Dunstone2006,Skelly2008}, similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \citep{Bouma2024}.   One interpretation for the
spectroscopic variability of these stars, and that of the analogous
transient \added{H$\alpha$ }absorption features
\citep{CollierCameron1989,CollierCameron1992,Cang2020}, comes from an
analogy to quiescent solar prominences, cool condensations of plasma
in the solar corona that can last days to weeks
\citep{VialEngvold2015}.  In the case of the Sun, these condensations
fall back to the photosphere because gravity is stronger than any
magnetic tension or centrifugal force capable of sustaining them.
However it has been understood at least since work by
\citet{Donati2000} that these ``prominence systems'' can be
longitudinally extended, forming a trapped ring of plasma.  While the
issue of why these previously known spectroscopic variables do not
show complex photometric variability remains open, plausible
explanations include that they are not in the required transiting
geometry, or that they lack the necessary source of opacity.  What we
have in the case of TIC\,141146667 is a system that finally exhibits
both sets of phenomena.

>>>>

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.99\textwidth]{f3.pdf}
  \vspace{-0.3cm}
  \captionsetup{labelformat=moviefmt,labelsep=colon}
	\caption{\textbf{Emission from plasma clumps orbiting
TIC\,141146667.}
  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{{\bf
  available here}},
  shows the spectral evolution over five hours.
  {\bf a,} TESS light curve from 5 February 2024 to 26
  February 2024 folded on the \periodhr\ hour period.  Black
  points are phase-averaged; gray are the raw data.
  {\bf b,} Keck/HIRES H$\alpha$ spectra acquired on 17 February 2024.
  The continuum is set to unity, and the darkest color is at twice the
  continuum to accentuate emission outside the line core ($|v/v_{\rm
  eq}|>1$, for the equatorial velocity $v_{\rm eq}$=130\,\kms).  While
  emission in the line core comes from the star's chromosphere, the
  sinusoidal emission features are most readily described by a warped
  plasma torus.
  {\bf c,} Individual epochs of Panel b, visible in the online movie.
  The dotted line shows the time-averaged spectrum, $f_{\langle t
  \rangle}$.
  {\bf d,} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf e,} As in Panel b, after subtracting $f_{\langle t
  \rangle}$.  The line core shows H$\alpha$ excesses and decrements
  advancing from the blue to red wings.
  The asymmetric color stretch mirrors the dynamic range of the data.
  {\bf f,} Individual epochs of Panel e, visible in the online
  movie.  \added{A separate version with overplotted best-fit sinusoids is also
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel_sinusoids.mp4}{{\bf
  available}}.}}
  \label{fig:spec}
\end{figure*}

@@

The H$\alpha$ line core\added{, visible in Panels e and f of
Figure~\ref{fig:spec}, }is more complex.  At $|\Delta v / v_{\rm
eq}|<1$, most observed H$\alpha$ photons come from the star's
chromosphere; \added{variability in the stellar emission line can be
caused by circumstellar material, bright regions, or dark regions
crossing the star's surface}.  Figure~\ref{fig:spec}e suggests that
\added{all three effects occur in TIC\,141146667}.  For instance, from
$\phi$=0-0.3, double-peaked emission is visible both on and off-limb;
this feature is circumstellar in origin.  However, the emission
feature that crosses the star from $\phi$=0.4-0.9 is brighter than the
circumstellar clumps, and its slow speed instead suggests it is
associated with a bright region on the star's surface.  Similarly, from
$\phi$=0.6-1.15 a 20\% deep absorption feature slowly crosses the
H$\alpha$ line profile.  This feature could be either a
chromospherically dark region (e.g.~a spot group), or an azimuthally
extended component of the circumstellar material.  The origin of the
other bright and dark streaks passing across the line core \added{in
Figure~\ref{fig:spec}e }is similarly ambiguous.

@@

Spectra of magnetically-active, rapidly rotating stars with a wide
range of masses have been previously observed to exhibit sinusoidal
time-varying Balmer emission
\citep{Donati2000,Townsend2005b,Dunstone2006,Skelly2008}, similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \citep{Bouma2024}.   One interpretation for the
spectroscopic variability of these stars, and that of the analogous
transient \added{H$\alpha$ }absorption features
\citep{CollierCameron1989,CollierCameron1992,Cang2020}, comes from an
analogy to quiescent solar prominences, cool condensations of plasma
in the solar corona that can last days to weeks
\citep{VialEngvold2015}.  In the case of the Sun, these condensations
fall back to the photosphere because gravity is stronger than any
magnetic tension or centrifugal force capable of sustaining them.
However, it has been understood at least since work by
\citet{Donati2000} that these ``prominence systems'' can be
longitudinally extended, forming a trapped ring of plasma.  While the
issue of why these previously known spectroscopic variables do not
show complex photometric variability remains open, plausible
explanations include that they are not in the required transiting
geometry, or that they lack the necessary source of opacity.  What we
have in the case of TIC\,141146667 is a system that finally exhibits
both sets of phenomena.
